---
title: "Total organic carbon - data exploration"
output: html_notebook
---

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
rm(list=ls())

if(!require("pacman")) install.packages("pacman"); library(pacman)
p_load(here, 
       CAST,
       corrplot,
       ggplot2)
```


## Load required data

```{r load_data}
rm_oc_sf <- read.csv(here("toc_model", "data", "interim", "rm_oc_sf.csv"))
rm_oc_ss <- read.csv(here("toc_model", "data", "interim", "rm_oc_ss.csv"))
predictors <- rast(here("toc_model", "data", "interim", "predictors.tif"))
```


# Data exploration

## Histogram of organic carbon content (weight-%)

```{r histogram_oc_stock, message=FALSE}
hist(rm_oc_sf$TOC, breaks = 28, main ="", xlab = "Surface TOC (weight-%)")
hist(rm_oc_ss$TOC, breaks = 28, main ="", xlab = "Subsurface TOC (weight-%)")

jpeg(filename = here("toc_model", "figures", "TOC_sf_histogram.jpg"), width = 15, height = 10, units = "cm", res = 300)
hist(rm_oc_sf$TOC, breaks = 28, main ="", xlab = "Surface TOC (weight-%)")
dev.off()

jpeg(filename = here("toc_model", "figures", "TOC_ss_histogram.jpg"), width = 15, height = 10, units = "cm", res = 300)
hist(rm_oc_ss$TOC, breaks = 28, main ="", xlab = "Subsurface TOC (weight-%)")
dev.off()
```


## Correlation plot

```{r corr_plot}
corrplot.mixed(cor(rm_oc_sf[, names(predictors)]), lower.col =  "black", tl.pos = "lt", number.cex = 0.6)

jpeg(filename = here("toc_model", "figures", "correlation_plot_sf.jpg"), width = 30, height = 30, units = "cm", res = 300)
corrplot.mixed(cor(rm_oc_sf[, names(predictors)]), lower.col =  "black", tl.pos = "lt", number.cex = 0.6)
dev.off()
```


## Environmental space

A visual check to what extent the samples cover the environmental space. This is useful as legacy data were used and no formal sampling design was applied in the analysis.

* Blue: Samples

* Grey: Environmental data (based on random subsample)

```{r env_space_sf}
smp <- as.data.frame(spatSample(x = predictors, size = nrow((rm_oc_sf)), method = "random", na.rm = TRUE))


for (i in names(predictors)) {
    
  print(ggplot() +
          geom_density(data = rm_oc_sf, aes(x=rm_oc_sf[,i]),colour="cornflowerblue",fill="cornflowerblue", alpha=0.1,linewidth=1) +
          geom_density(data = smp, aes(x=smp[,i]), colour="grey",fill="grey", alpha=0.1, linewidth=1) +
          scale_x_continuous(name = names(rm_oc_sf[i])))
        
}
```




