---
title: "Fishing data"
output: html_notebook
---

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
rm(list=ls())

if(!require("pacman")) install.packages("pacman"); library(pacman)
p_load(here,
       sf,
       terra)
```


## Define variables

Available gear types: Beam, Dredge, Otter and Seine. Dredge is not included in this analysis as there is no footprint in Norwegian waters.
Years: 2009 to 2020
Penetration depths: sar (0 - 2 cm) and subsar (2 - 5 cm). The analysis is only looking at surface SAR.

```{r variables}
gear <- c("Beam", "Otter", "Seine")
year <- c(2009:2020)
var <- "sar"
```



## Download data from figshare

ICES (2021). Data for OSPAR request on the production of spatial data layers of fishing intensity/pressure. Data Outputs. Dataset. https://doi.org/10.17895/ices.data.8294
Download the data and save it in fishing_data/data/raw.

```{r download}
unzip(zipfile = here("fishing_data", "data", "raw", "ICES.2021.OSPAR_production_of_spatial_fishing_pressure_data_layers.zip"), files = "shapefiles.zip", exdir = here("fishing_data", "data", "raw"))

# get names of files to unzip
files <- unzip(zipfile = here("fishing_data", "data", "raw", "shapefiles.zip"), list = TRUE)
files <- files[,-c(2,3)]
files <- subset(files, grepl(gear[1], files) | grepl(gear[2], files) | grepl(gear[3], files))

unzip(zipfile = here("fishing_data", "data", "raw", "shapefiles.zip"), files = files, exdir = here("fishing_data", "data", "raw"))

file.remove(here("fishing_data", "data", "raw", "shapefiles.zip"))
file.remove(here("fishing_data", "data", "raw", "ICES.2021.OSPAR_production_of_spatial_fishing_pressure_data_layers.zip"))
```


## Create templates

```{r templates}
t_r <- rast(nrows = 535, ncols = 1100, xmin = -15, xmax = 40, ymin = 55, ymax = 81.75, crs = "EPSG:4326", vals = 1)
t_v <- st_as_sf(as.polygons(t_r, trunc = FALSE, dissolve = FALSE))
```


# Create raster data from shapefiles

This step is time-consuming.

```{r create_rasters}
for (i in gear) {
  for (j in year) {
    f <- read_sf(paste0(here("fishing_data", "data", "raw"), "/", i, "-", j, ".shp"))
    head(f)
    
    f <- centroids(vect(f), inside = TRUE)
    
    join <- st_join(t_v, st_as_sf(f))
    r <- rasterize(vect(join), t_r, field = var)
    
    writeRaster(r, paste0(here("fishing_data", "data", "interim"), "/", var, "_", i, "_", j, ".tif"), overwrite = TRUE)
  }
}
```


# Calculate statistics

## Beam trawls

### Surface SAR

```{r sar_beam}
sar_Beam_2009_2020 <- rast()

sar_Beam <- rast(list.files(path = here("fishing_data", "data", "interim"), pattern = 'sar_Beam', full.names = TRUE))

#assign NA values to 0
sar_Beam[is.na(sar_Beam)] <- 0

#assign values of -9 to NA
sar_Beam[sar_Beam == -9] <- NA

#calculate annual statistics
sar_Beam_2009_2020$SARmin <- min(sar_Beam, na.rm = TRUE)
sar_Beam_2009_2020$SARmean <- mean(sar_Beam, na.rm = TRUE)
sar_Beam_2009_2020$SARmax <- max(sar_Beam, na.rm = TRUE)
sar_Beam_2009_2020$SARsum <- sum(sar_Beam, na.rm = TRUE)
sar_Beam_2009_2020$SARrange <- range(sar_Beam, na.rm = TRUE)

plot(sar_Beam_2009_2020)
```


## Otter trawls

### Surface SAR

```{r sar_otter}
sar_Otter_2009_2020 <- rast()

sar_Otter <- rast(list.files(path = here("fishing_data", "data", "interim"), pattern = 'sar_Otter', full.names = TRUE))

#assign NA values to 0
sar_Otter[is.na(sar_Otter)] <- 0

#assign values of -9 to NA
sar_Otter[sar_Otter == -9] <- NA

#calculate annual statistics
sar_Otter_2009_2020$SARmin <- min(sar_Otter, na.rm = TRUE)
sar_Otter_2009_2020$SARmean <- mean(sar_Otter, na.rm = TRUE)
sar_Otter_2009_2020$SARmax <- max(sar_Otter, na.rm = TRUE)
sar_Otter_2009_2020$SARsum <- sum(sar_Otter, na.rm = TRUE)
sar_Otter_2009_2020$SARrange <- range(sar_Otter, na.rm = TRUE)

plot(sar_Otter_2009_2020)
```


## Seines

### Surface SAR

```{r sar_seines}
sar_Seine_2009_2020 <- rast()

sar_Seine <- rast(list.files(path = here("fishing_data", "data", "interim"), pattern = 'sar_Seine', full.names = TRUE))

#assign NA values to 0
sar_Seine[is.na(sar_Seine)] <- 0

#assign values of -9 to NA
sar_Seine[sar_Seine == -9] <- NA

#calculate annual statistics
sar_Seine_2009_2020$SARmin <- min(sar_Seine, na.rm = TRUE)
sar_Seine_2009_2020$SARmean <- mean(sar_Seine, na.rm = TRUE)
sar_Seine_2009_2020$SARmax <- max(sar_Seine, na.rm = TRUE)
sar_Seine_2009_2020$SARsum <- sum(sar_Seine, na.rm = TRUE)
sar_Seine_2009_2020$SARrange <- range(sar_Seine, na.rm = TRUE)

plot(sar_Seine_2009_2020)
```


## Total

### Surface SAR

```{r sar_total}
sar_total_2009_2020 <- sar_Beam_2009_2020 + sar_Otter_2009_2020 + sar_Seine_2009_2020
```


# Output raster

```{r write_raster}
writeRaster(sar_Beam_2009_2020$SARmean, here("fishing_data", "data", "final", paste0(var, "_mean_Beam_2009to2020.tif")), overwrite = TRUE)
writeRaster(sar_Otter_2009_2020$SARmean, here("fishing_data", "data", "final", paste0(var, "_mean_Otter_2009to2020.tif")), overwrite = TRUE)
writeRaster(sar_Seine_2009_2020$SARmean, here("fishing_data", "data", "final", paste0(var, "_mean_Seine_2009to2020.tif")), overwrite = TRUE)
writeRaster(sar_total_2009_2020$SARmean, here("fishing_data", "data", "final", paste0(var, "_mean_total_2009to2020.tif")), overwrite = TRUE)
```



